Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-07-29 jso.qmd
add definition for squint angle

```{=tex}
\begin{defn}[projection distance]\label{def:proj-dist}
Let $A \in \mathbf{R}^{p \times d}$ be the d-dimensional projection basis and
$A^*$ be the optimal basis that achieves the maximum index value for a given
PP problem defined by the shape to find, data dimension,
and the index function. The projection distance $d(A, A^*)$ between $A$ 
and $A^*$ is defined as 
$d(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\  \rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}
```
We now focus our attention to how the index value $f(XA)$ changes with
respect to the projection distance $d(A, A^*)$, over the course of the
JSO.

<!-- Squintability is defined as how rapidly and significantly the index value, $f(XA)$ response to the decrease of projection distance to optimal, $d(A, A^*)$. -->

```{=tex}
\begin{defn}[squintability]\label{def:squintability}
Let $g: \mathbb{R} \mapsto  \mathbb{R}$ be a decreasing function that maps the projection distance, $d = d(A, A^*)$, 
to the index value $f(XA)$, such that $g(d(A, A^*)) = f(XA)$. The squintability of the index function $f(\cdot)$, is defined as 
$$\varsigma(f) = -c \times \max_{d} g'(d) \times \arg \max_{d} g'(d)$$ 
where $c$ is a constant scaling factor, $-\max_d g'(d)$ represents the 
largest gradient of $-g$ and $\arg \max_{d} g'(d)$ represents the projection distance at which this largest gradient is attained.

>>>>

```{=tex}
\begin{defn}[projection distance]\label{def:proj-dist}
Recall that $A \in \mathbf{R}^{p \times d}$ is a $d$-dimensional orthonormal matrix. Let $A^*$ be the optimal matrix that achieves the maximum index value for a given data. The projection distance $d(A, A^*)$ between $A$ 
and $A^*$ is defined as 
$d(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\  \rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}
```

We now focus our attention to how the index value $f(XA)$ changes with
respect to the projection distance $d(A, A^*)$, over the course of the
JSO.

```{=tex}
\begin{defn}[squint angle]\label{def:squint-angle}
Let $A$ and $B$ be two $d$-dimensional orthonormal matrices in $\mathbb{R}^p$. The squint angle $\theta$ between the subspace spanned by $A$ and $B$ is defined as the smallest principal angle between these subspaces: $\theta = \min_{i \in \{1, \cdots, d\}} \arccos(\tau_i)$, where $\tau_i$ are the singular values of the matrix $M = A^T B$ obtained from its singular value decomposition.

```{=tex}
\begin{defn}[squintability]\label{def:squintability}
Let $g: \mathbb{R} \mapsto  \mathbb{R}$ be a decreasing function that maps the projection distance $d(A, A^*)$ to the index value $f(XA)$, such that $g(d(A, A^*)) = f(XA)$. For brevity, denote $g(d)$ as $g(d(A, A^*))$. The squintability of an index function $f(XA)$ is defined as

$$\varsigma(f) = -c \times \max_{d} g'(d) \times \arg \max_{d} g'(d) $$

where $c$ is a constant scaling factor, $-\max_d g'(d)$ represents the 
largest gradient of $-g$ and $\arg \max_{d} g'(d)$ represents the projection distance at which this largest gradient is attained.

### 2024-07-29 jso.tex
add definition for squint angle

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, toprule=.15mm, bottomtitle=1mm, toptitle=1mm, arc=.35mm, coltitle=black, left=2mm, breakable, title={Algorithm: Jellyfish Optimizer Pseudo Code}, colframe=quarto-callout-note-color-frame, rightrule=.15mm, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]

@@

\begin{defn}[projection distance]\label{def:proj-dist}
Let $A \in \mathbf{R}^{p \times d}$ be the d-dimensional projection basis and
$A^*$ be the optimal basis that achieves the maximum index value for a given
PP problem defined by the shape to find, data dimension,
and the index function. The projection distance $d(A, A^*)$ between $A$ 
and $A^*$ is defined as 
$d(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\  \rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}

@@

\begin{defn}[squintability]\label{def:squintability}
Let $g: \mathbb{R} \mapsto  \mathbb{R}$ be a decreasing function that maps the projection distance, $d = d(A, A^*)$, 
to the index value $f(XA)$, such that $g(d(A, A^*)) = f(XA)$. The squintability of the index function $f(\cdot)$, is defined as 
$$\varsigma(f) = -c \times \max_{d} g'(d) \times \arg \max_{d} g'(d)$$ 
where $c$ is a constant scaling factor, $-\max_d g'(d)$ represents the 
largest gradient of $-g$ and $\arg \max_{d} g'(d)$ represents the projection distance at which this largest gradient is attained.

@@

It is expected that for a PP index with high squint angle, the
optimization (\ref{eq-optimization}) should make substantial progress
early on. Conversely, for a PP index with low squint angle, it might
take a long while for the optimization to make substantial progress, as
the candidate projections would need to be very close to the optimal one
for the structure of the index function to be visible enough to be
amenable to efficient optimization. This observation suggests that the
extreme values of \(f'\) (the ones for which \(f''=0\), assuming that
\(f\) is twice differentiable), and the projection distances for which
these values are attained, are crucial in the mathematical definition of
squintability. Noticing that \(f\) is expected to be a decreasing
function, the squintability metric is proposed as follows:

@@

\(-\min_{x}f'(x)\) represents the largest value of the gradient of
\(-f\) and \(\underset{x}{\arg\min}f'(x)\) represents the projection
distance at which this largest gradient is attained. It is expected that
these two values should be both high in the case of high squintability
(fast increase in \(f\) early on), and both low in the case of low
squintability (any substantial increase in \(f\) happens very late,
close to the optimal angle). This suggest that their product
(\ref{eq-squintability}) should provide a sensible measure of
squintability. The multiplicative constant \(4\), which can be deemed
arbitrary, does not change the interpretation of the squintability
metric \(\varsigma\); it is here to adjust the range of values of
\(\varsigma\) and simplify the explicit formula for \(\varsigma\)
obtained later on.

>>>>

\begin{tcolorbox}[enhanced jigsaw, colback=white, opacityback=0, titlerule=0mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, rightrule=.15mm, coltitle=black, breakable, left=2mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, toprule=.15mm]

@@

\begin{defn}[projection distance]\label{def:proj-dist}
Recall that $A \in \mathbf{R}^{p \times d}$ is a $d$-dimensional orthonormal matrix. Let $A^*$ be the optimal matrix that achieves the maximum index value for a given data. The projection distance $d(A, A^*)$ between $A$ 
and $A^*$ is defined as 
$d(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\  \rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}

@@

\begin{defn}[squint angle]\label{def:squint-angle}
Let $A$ and $B$ be two $d$-dimensional orthonormal matrices in $\mathbb{R}^p$. The squint angle $\theta$ between the subspace spanned by $A$ and $B$ is defined as the smallest principal angle between these subspaces: $\theta = \min_{i \in \{1, \cdots, d\}} \arccos(\tau_i)$, where $\tau_i$ are the singular values of the matrix $M = A^T B$ obtained from its singular value decomposition.

\begin{defn}[squintability]\label{def:squintability}
Let $g: \mathbb{R} \mapsto  \mathbb{R}$ be a decreasing function that maps the projection distance $d(A, A^*)$ to the index value $f(XA)$, such that $g(d(A, A^*)) = f(XA)$. For brevity, denote $g(d)$ as $g(d(A, A^*))$. The squintability of an index function $f(XA)$ is defined as

$$\varsigma(f) = -c \times \max_{d} g'(d) \times \arg \max_{d} g'(d) $$

where $c$ is a constant scaling factor, $-\max_d g'(d)$ represents the 
largest gradient of $-g$ and $\arg \max_{d} g'(d)$ represents the projection distance at which this largest gradient is attained.

@@

It is expected that for a PP index with high squint angle, the
optimization (\ref{eq-optimization}) should make substantial progress
early on. Conversely, for a PP index with low squint angle, it might
take a long while for the optimization to make substantial progress, as
the candidate projections would need to be very close to the optimal one
for the structure of the index function to be visible enough to be
amenable to efficient optimization. This observation suggests that the
extreme values of \(f'\) (the ones for which \(f''=0\), assuming that
\(f\) is twice differentiable), and the projection distances for which
these values are attained, are crucial in the mathematical definition of
squintability.

Noticing that \(f\) is expected to be a decreasing function, the
squintability metric is proposed as follows:

@@

\(-\min_{x}f'(x)\) represents the largest value of the gradient of
\(-f\) and \(\underset{x}{\arg\min}f'(x)\) represents the projection
distance at which this largest gradient is attained.

It is expected that these two values should be both high in the case of
high squintability (fast increase in \(f\) early on), and both low in
the case of low squintability (any substantial increase in \(f\) happens
very late, close to the optimal angle). This suggest that their product
(\ref{eq-squintability}) should provide a sensible measure of
squintability.

The multiplicative constant \(4\), which can be deemed arbitrary, does
not change the interpretation of the squintability metric \(\varsigma\);
it is here to adjust the range of values of \(\varsigma\) and simplify
the explicit formula for \(\varsigma\) obtained later on.
